---
title: "BD : Nettoyage de NYCFlights"


execute:
  eval: false
  echo: true
---

```{.r, include=FALSE}
library(DBI)
require(dbplyr)
require(tidyverse)

db <- dbConnect(RPostgreSQL::PostgreSQL(), 
                user="boucheron",
                password=rstudioapi::askForPassword("Database password"),
                host="localhost",
                port=5434,
                dbname = "bd_2023")

knitr::opts_chunk$set(connection = "db")
```



```{sql, connection=db}
SELECT * FROM nycflights13.airlines
```


```{sql, connection=db}
SELECT * 
FROM nycflights13.planes 
LIMIT 10
```

# Exploration des valeurs manquantes 

---

```{sql, connection=db}
SELECT *  
FROM nycflights13.flights 
LIMIT 10 ;
```

```{sql, connection=db}
SELECT COUNT(*)  
FROM nycflights13.flights  f1 JOIN nycflights13.flights  f2
ON (f1.flight=f2.flight AND f1.carrier = f2.carrier AND f1.origin<> f2.origin) ;
```

---

```{sql, connection=db}
WITH R AS (
  SELECT f1.flight, f1.carrier, 
    COUNT(DISTINCT f1.dest)  AS n_dest, 
    COUNT(DISTINCT f1.origin) AS n_origin
  FROM nycflights13.flights  f1 
  GROUP BY f1.flight, f1.carrier 
  HAVING COUNT(DISTINCT f1.dest) > 1 OR COUNT(DISTINCT f1.origin) > 1
)

SELECT R.carrier, COUNT(*) AS n
FROM R 
GROUP BY R.carrier 
ORDER BY n DESC;
```

---

```{sql, connection=db}
WITH R AS (
  SELECT f1.carrier, f1.origin, f1.dest, f1.sched_dep_time, 
    COUNT(DISTINCT f1.flight)
  FROM nycflights13.flights  f1 
  GROUP BY f1.carrier, f1.origin, f1.dest, f1.sched_dep_time 
  HAVING COUNT(DISTINCT f1.flight) >  1
) 

SELECT carrier, COUNT(*)
FROM R 
GROUP BY carrier ;
```

---

```{sql, connection=db}
SELECT *
FROM nycflights13.flights  f1  ;
```

---

```{r}

```

```{r}
flights2 <- tbl(db, in_schema('nycflights13','flights'))
```

---

```{r}
sk_flights <- flights2 |> 
  skimr::skim() 
```

Colonnes avec valeurs manquantes 
```{r}
sk_flights |> 
  filter(n_missing >0 )
```

Dépendances fonctionnelles 


```{r}
require(lubridate)

flights2 |> 
  filter(lubridate::year(time_hour) != year | 
         lubridate::month(time_hour) != month| 
         lubridate::day(time_hour) != day |
         lubridate::hour(time_hour) != hour |
         sched_dep_time / 100L  != hour |
         sched_dep_time %% 100L  != minute ) |> 
  head()
```

```{r}
flights2 |> 
  group_by(origin, dest) |> 
  summarise(n_dist = n_distinct(distance), .groups="drop") |> 
  filter(n_dist > 1) |>  
  head()
```


```{r}
flights2 |> 
  filter(origin %in% c("EWR", "JFK"), dest=="EGE") |> 
  select(carrier, origin, dest, distance) |> 
  distinct() 
```
```{r}
flights2 |> 
  group_by(carrier, origin, wday(time_hour), sched_dep_time) |> 
  summarise(n_flights= n_distinct(flight), n_dest=n_distinct(dest), .groups = "drop") |> 
  filter( n_dest >1) |> 
  group_by(carrier) |> 
  summarise(n())
```


[https://en.wikipedia.org/wiki/Flight_number](https://en.wikipedia.org/wiki/Flight_number)


```{r}
res <- dbSendQuery(db, "
    WITH R AS (
SELECT f1.carrier, f1.origin, f1.dest, f1.sched_dep_time, COUNT(DISTINCT f1.flight)
FROM nycflights13.flights  f1 
GROUP BY f1.carrier, f1.origin, f1.dest, f1.sched_dep_time 
HAVING COUNT(DISTINCT f1.flight) >  1) 
SELECT carrier, COUNT(*)
FROM R 
GROUP BY carrier ;
")
```

```{r}
# dbListResults(db)
dbFetch(res, n=10)
dbClearResult(res)
#dbDisconnect(db)
```

# Understanding `planes`

```{r}
planes2 <- tbl(db, in_schema('nycflights13','planes'))
```

Conjecture : `model` et `year`  déterminent d'autres attributs, comme `manufacturer`, `engines`, `engine`, `type`. 

Pour une vingtaine de couples `(model, year)` ne détermine pas tous ces attributs. 

Pour un couple, `(model, year)` ne détermine ni `manufacturer`, ni `engine`, ni `engines`. 
Pour neuf couples,  `(model, year)` ne détermine ni `manufacturer` ni `engine`. 
Pour un couple, `(model, year)` détermine 
```{r}
planes2 |> 
  group_by(model, year) |> 
  summarize(n_distinct(manufacturer),  n_distinct(engine), n_distinct(engines), n_distinct(type), .groups = "drop") |> 
  filter(`n_distinct(manufacturer)`>1 |  `n_distinct(engine)`>1) |> 
  arrange(desc(`n_distinct(engine)`))
```

```{sql, connection=db}
SELECT model, year, 
       array_agg(DISTINCT manufacturer) as manu, 
       array_agg(DISTINCT engine) AS engie, 
       array_agg(DISTINCT engines) AS engies
FROM nycflights13.planes pl 
GROUP BY model, year
HAVING COUNT(DISTINCT manufacturer) >1 OR 
       COUNT(DISTINCT engine) > 1 ;
```

```{r}
planes2 |> 
  filter(substr(model, 1, 5) =="A330-") |> 
  select(model, manufacturer, engines, engine)

planes2 |> 
  filter(engine=="Turbo-jet")

planes2 |> 
  filter(substr(manufacturer, 1, 6) =="MCDONN") |> 
  distinct(year, manufacturer,engine, model) |> 
  arrange(year)

```
The last  tow lines look erroneous. `A330-xxx` have two engines of type `Turbo-fan`. The manufacturer   should be named `AIRBUS` (`AIRBUS INDUSTRIE` was the most frequent denomination until 2002).


> What is the difference between a turbojet and a turbofan? The main difference between a turbofan and a turbojet is that all the air goes into the engine core (compressor, combustion chamber, turbine) in a turbojet. A turbofan uses the main elements of a turbojet but has a large fan at the front of the engine.

```{sql, connection=db}
CREATE TABLE nyc_flights.plane_models AS 
SELECT DISTINCT model, year, manufacturer, engine, engines, type 
FROM nycflights13.planes pl ;
```

```{sql,  connection=db}
UPDATE nyc_flights.plane_models
SET  manufacturer='AIRBUS', engine='Turbo-fan', engines=2
WHERE model='A330-223' ;
```

```{r}
planes3 <- tbl(db, in_schema('nyc_flights','plane_models'))
```

```{r}
planes3 |> 
  group_by(model, year) |> 
  summarize(n_distinct(manufacturer),  
            n_distinct(engine), 
            n_distinct(engines), 
            n_distinct(type), 
            .groups = "drop") |> 
  filter(`n_distinct(manufacturer)`>1 |  
         `n_distinct(engine)`>1 ) |> 
  arrange(desc(`n_distinct(engine)`))
```

```{sql, connection=db}
UPDATE nyc_flights.plane_models
SET  manufacturer='MCDONNELL DOUGLAS'
WHERE manufacturer LIKE 'MCDONNELL DOUGLAS %' ;
```

```{r}
planes3 |> 
  filter(substr(manufacturer, 1, 6) =="MCDONN") |> 
  distinct(year, manufacturer,engine, model) |> 
  arrange(year)
```


# Fin